Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-24 paper/section-instantiations.tex
finally proved iso

\begin{example}[Geometry: wlog]\label{ex:simpletrans}
We prove that
\[\DeltaWithB\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to\tyPrim{vec}{B,t}) \cong
\tyPrim{vec}{B,0}.
\]
Let 
\begin{align*}
X & \isDefinedAs\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to\tyPrim{vec}{B,t}
\quad\text{and}\quad Y \isDefinedAs\tyPrim{vec}{B,0}\\
I & \isDefinedAs\lambda f\mathord:X.f\,[0]\,(0\,[B])\\
J & \isDefinedAs\lambda v\mathord:Y.\Lambda t\mathord:\SynTransl{2}.\lambda w\mathord:\tyPrim{vec}{B,t}.v +[B,0,t] w
\end{align*}
Unfolding definitions gives that $\tmSem{I(J(v))}(\opsEnv,\vec{v})$ = $\vec{v}+(0,0)$
for any $\vec{v}\in\mathbb R ^ 2$.  Because $(0,0)$ is the identity for vector
addition it follows that $\ctxteq{\DeltaWithB}{v\mathord:Y}{I(J(v))}{v}Y$.
To show $\ctxteq{\DeltaWithB}{f\mathord:X}{J(I(f))}{f}{X}$ we appeal to
\autoref{thm:soundness} and reason using the relational semantics.  It
suffices to show $\semeq{\DeltaWithB}{f\mathord:X}{J(I(f))}{f}{X}$; that
is, given $\rho\in\semSort\DeltaWithB$, for any $f$ and $f'$ with
$(f,f')\in\rsem X\rho$, it is the case that 
$(\tmSem{J(I(f))}(\opsEnv,f), f')\in\rsem X\rho$.
Expanding the premise, we have
\begin{eqnarray}\label{eqn:transpremise}
\forall \vec{t}\in \mathbb R ^ 2,
\forall \vec{v}\in \mathbb R ^ 2,
f'(\rho(B)\vec{v} + \vec{t}) = \rho(B)(f(\vec{v})) + \vec{t}
\end{eqnarray}
Expanding the conclusion, we have to show that for any
$\vec{t_0}\in\mathbb R^2$ and $\vec{v_0}\in\mathbb R^2$, it is the case that
$f'(\rho(B)\vec{v_0} + \vec{t_0}) = \rho(B)(\vec{v_0} + f(0)) + \vec{t_0}$.
By instantiating \autoref{eqn:transpremise} thrice, with~$\vec{v}=\vec{t}=(0,0)$, with~$\vec{v}=\vec{v_0}$ and~$\vec{t}=\vec{t_0}$,
and with (crucially)~$\vec{v}=\vec{v_0}$ and~$\vec{t}=-\rho(B)\vec{v_0}$, and applying a little algebra, we obtain
the desired result.
\qed
\end{example}

@@

\newcommand{\transup}[3]{\mathord\uparrow_{#1}^{#2\mathord:#3}}
\newcommand{\transdn}[3]{\mathord\downarrow_{#1}^{#2\mathord:#3}}

\begin{lemma}\label{lem:transiso}
Let $A$ be a quantifier-free type in which every
occurrence of $\tyPrimNm{vec}$ is of the form $\tyPrim{vec}{B,e}$ for some $e$. Then
\[
\DeltaWithB\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to A)
\cong A[0/t]
\]
\end{lemma}
\begin{proof}
Let 
\begin{align*}
X & \isDefinedAs \forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to A
& 
Y & \isDefinedAs A[0/t]\\
I & \isDefinedAs \lambda f\mathord: X. f\ [0]\ (0\ [B])
& 
J & \isDefinedAs \lambda y\mathord: Y. \Lambda t.\lambda v\mathord:\tyPrim{vec}{B,t}. \transup A t v (y)
\end{align*}
where terms 
\begin{align*}
&
\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transup A t v : A[0/t] \to A
\\\text{and } &
\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transdn A t v : A\to A[0/t]
\end{align*}
are defined by induction over the syntax of $A$. For brevity we omit types on binders
and indices on primitives; also note that the omitted cases for $\transup A t v$ 
have definition symmetric to $\transdn A t v$:
\begin{align*}
\transdn{\tyPrimNm{unit}} t v &= \lambda x. x
\qquad 
\transdn{\tyPrim{real}{e}} t v = \lambda x. x
\\
\transdn{A_1\times A_2}t v &= \lambda p. (\transdn{A_1}t v(\pi_1 p),\transdn{A_2}t v(\pi_2 p))
\\
\transdn{A_1 + A_2} t v &= \lambda x.
\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn {A_1}t v y); \textrm{inr}\ y.\textrm{inr}(\transdn {A_2} t v y)
\\ 
\transdn{A_1\to A_2} t v &= \lambda f. \lambda x. \transdn{A_2} t v (f(\transup {A_1} t v x))
\end{align*}
\begin{align*}
\transdn{\tyPrim{vec}{B,e}} t v &= \lambda x. \transdn{e}  t v (x) 
&
\transup{\tyPrim{vec}{B,e}} t v &= \lambda x. \transup{e} t v (x)
\\
\transdn{0} t v &= \lambda w.w
&
\transup{0} t v &= \lambda w.w
\\ 
\transdn{e_1+e_2} t v &= \lambda w.\transdn{e_1} t v(\transdn{e_2} t v(w))
&
\transup{e_1+e_2} t v &= \lambda w.\transup{e_1} t v(\transup{e_2} t v(w))
\\
\transdn{-e} t v &= \lambda w.\transup{e} t v (w)
&
\transup{-e} t v &= \lambda w.\transdn{e} t v (w)
\\
\transdn{t} t v &= \lambda w.w + v 
&
\transup{t} t v &= \lambda w.w + (\mathrm{negate}\, v)
\\
\transdn{t'} t v &= \lambda w.w\quad(t'\neq t)
&
\transup{t'} t v &= \lambda w.w\quad(t'\neq t)
\end{align*}
To show $I(J(y))$ is contextually equivalent to $y$ is an easy
induction on $A$ using the fact that the vector $(0,0)$
is an identity for vector addition. To show $J(I(x))$ is contextually
equivalent to $x$ we first prove (by induction on $A$) 
that for any $\rho\in\semSort{\DeltaWithB}$, and any $\vec v\in\mathbb R^2$, 
the graph of the function given by $\tmSem{\transdn A t v}(\opsEnv,\vec v)$
is just  $\rsem A (\rho,\vec v)$, and conversely, the graph of $\tmSem{\transup A t v}(\opsEnv,\vec v)$
is ${\rsem A (\rho,\vec v)}^{\mathrm{op}}$.
\end{proof}

\exref{ex:type-iso} and \exref{ex:simpletrans} are special cases of
this isomorphism. Another instance is the type of vector addition:
\begin{align*}
\forall B\mathord:\SynGL{2}, t_1, t_2\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_2}\to\tyPrim{vec}{B,t_1\mathord+t_2}\\
\cong\quad\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_1}
\quad (\text{by~\lemref{lem:transiso}})\\
\cong\quad\forall B\mathord:\SynGL{2}.\tyPrim{vec}{B,0} \quad (\text{by~\lemref{lem:transiso}})\\
\cong\quad\tyUnit\quad(\text{by abstraction theorem})
\end{align*}
In other words, the vector addition operation is the only inhabitant of its type!
\paragraph{Scalings}
For the group $\GL{1}$ of scalings we can treat
a real-valued argument to a function as a unit-of-measure with
which to scale the result. 
%The small fly in the ointment is that the
Although the argument might be zero, and this cannot be used for scaling,
%Nevertheless, 
we still can obtain the following slightly more complicated
isomorphism:
\begin{eqnarray}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A &
\cong & [1/s]\times \forall s\mathord:\SynGL{1}.A\label{eqn:scaleiso}
\end{eqnarray}
The type of $\mathrm{areaCircle}$ from the introduction is one instance:
\begin{align*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s^2}\\ 
\cong\quad\tyPrim{real}{1}\times(\forall
s\mathord:\SynGL{1}.\tyPrim{real}{s^2}) \quad(\text{by }\ref{eqn:scaleiso})\\ 
\cong\quad\tyPrim{real}{1}\times \tyUnit \quad (\text{by abstraction theorem})\\
\cong\quad\tyPrim{real}{1}\quad (\text{trivially})
\end{align*}
%where %the isomorphism 
%The abstraction theorem is used to prove
%$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s^2}\cong \tyUnit$.
% is proved using
%the abstraction theorem.
%\begin{enumerate}
%\item Two cases: when we have a total ``division'' operation, and when
%  we don't.
%\item Type isomorphisms, based on Smith Normal Form
% \item Non-definability results, based on the non-compositional
%   relational environments
% \item Alteration of the non-definability results in the light of
%   addition of a square root operation
% \item Mention the variation we can obtain by changing some types,
%   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
%\end{enumerate}

>>>>

\begin{example}[Geometry: wlog]\label{ex:simpletrans}
We prove that
\[\DeltaWithB\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to\tyPrim{vec}{B,t}) \cong
\tyPrim{vec}{B,0}.
\]
Let 
\begin{align*}
X & \isDefinedAs\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to\tyPrim{vec}{B,t}
\quad\text{and}\quad Y \isDefinedAs\tyPrim{vec}{B,0}\\
I & \isDefinedAs\lambda f\mathord:X.f\,[0]\,(0\,[B])\\
J & \isDefinedAs\lambda v\mathord:Y.\Lambda t\mathord:\SynTransl{2}.\lambda w\mathord:\tyPrim{vec}{B,t}.v +[B,0,t] w
\end{align*}
Unfolding definitions gives that $\tmSem{I(J(v))}(\opsEnv,\vec{v})$ = $\vec{v}+\vec{0}$
for any $\vec{v}\in\mathbb R ^ 2$.  Because $\vec{0}$ is the identity for vector
addition it follows that $\ctxteq{\DeltaWithB}{v\mathord:Y}{I(J(v))}{v}Y$.
To show $\ctxteq{\DeltaWithB}{f\mathord:X}{J(I(f))}{f}{X}$ we appeal to
\autoref{thm:soundness} and reason using the relational semantics.  It
suffices to show $\semeq{\DeltaWithB}{f\mathord:X}{J(I(f))}{f}{X}$; that
is, given $\rho\in\semSort\DeltaWithB$, for any $f$ and $f'$ with
$(f,f')\in\rsem X\rho$, it is the case that 
$(\tmSem{J(I(f))}(\opsEnv,f), f')\in\rsem X\rho$.
Expanding the premise, we have
\begin{eqnarray}\label{eqn:transpremise}
\forall \vec{t}\in \mathbb R ^ 2,
\forall \vec{v}\in \mathbb R ^ 2,
f'(\rho(B)\vec{v} + \vec{t}) = \rho(B)(f(\vec{v})) + \vec{t}
\end{eqnarray}
Expanding the conclusion, we have to show that for any
$\vec{t_0}\in\mathbb R^2$ and $\vec{v_0}\in\mathbb R^2$, it is the case that
$f'(\rho(B)\vec{v_0} + \vec{t_0}) = \rho(B)(\vec{v_0} + f(0)) + \vec{t_0}$.
By instantiating \autoref{eqn:transpremise} thrice, with~$\vec{v}=\vec{t}=\vec{0}$, with~$\vec{v}=\vec{v_0}$ and~$\vec{t}=\vec{t_0}$,
and with (crucially)~$\vec{v}=\vec{v_0}$ and~$\vec{t}=-\rho(B)\vec{v_0}$, and applying a little algebra, we obtain
the desired result.
\qed
\end{example}

@@

\newcommand{\transup}[3]{\mathord\uparrow_{#1}^{#2\mathord:#3}}
\newcommand{\transdn}[3]{\mathord\downarrow_{#1}^{#2\mathord:#3}}
\newcommand{\transtrans}[3]{\mathrm{tr}_{#1}^{#2\mathord:#3}}

\begin{lemma}\label{lem:transiso}
Suppose that $\DeltaWithB,t\mathord:\SynTransl{2}\vdash A \isType$, 
$A$ contains no quantifiers and every occurrence of $\tyPrimNm{vec}$ is of the form $\tyPrim{vec}{B,?}$. Then
\[
\DeltaWithB\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to A)
\cong A[0/t]
\]
\end{lemma}
\begin{proof}
Let 
\begin{align*}
X & \isDefinedAs \forall t\mathord:\SynTransl{2}.\tyPrim{vec}{B,t}\to A
\quad\text{and}\quad Y \isDefinedAs A[0/t]\\
I & \isDefinedAs \lambda f\mathord: X. f\ [0]\ (0\ [B])\\
J & \isDefinedAs \lambda y\mathord: Y. \Lambda t.\lambda v\mathord:\tyPrim{vec}{B,t}. \transup A t v (y)
\end{align*}
where terms 
\begin{align*}
&\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transup A t v : A[0/t] \to A\\
&\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transdn A t v : A\to A[0/t]\\
&\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transup e t v : \tyPrim{vec}{B,e[0/t]}\to\tyPrim{vec}{B,e}\\
&\DeltaWithB,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{B,t}\vdash \transdn e t v : \tyPrim{vec}{B,e}\to\tyPrim{vec}{B,e[0/t]}
\end{align*}
are defined by induction over the structure of $A$ and $e$. For brevity we omit types on binders;
also note that the omitted cases for $\transdn A t v$ 
have definition symmetric to $\transup A t v$:
\begin{align*}
\transup{\tyPrimNm{unit}} t v &= \lambda x. x
\qquad 
\transup{\tyPrim{real}{e}} t v = \lambda x. x
\\
\transup{A_1\times A_2}t v &= \lambda p. (\transup{A_1}t v(\pi_1 p),\transup{A_2}t v(\pi_2 p))
\\
\transup{A_1 + A_2} t v &= \lambda x.
\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transup {A_1}t v y); \textrm{inr}\ y.\textrm{inr}(\transup {A_2} t v y)
\\ 
\transup{A_1\to A_2} t v &= \lambda f. \lambda x. \transup{A_2} t v (f(\transdn {A_1} t v x))
\end{align*}
\begin{align*}
\transup{\tyPrim{vec}{B,e}} t v &= \lambda x. \transdn{e}  t v (x) 
&
\transdn{\tyPrim{vec}{B,e}} t v &= \lambda x. \transup{e} t v (x)
\\
\transup{0} t v &= \lambda w.w
&
\transdn{0} t v &= \lambda w.w
\\ 
\transup{e_1+e_2} t v &= \lambda w.\transup{e_1} t v(\transup{e_2} t v(w))
&
\transdn{e_1+e_2} t v &= \lambda w.\transdn{e_1} t v(\transdn{e_2} t v(w))
\\
\transup{-e} t v &= \lambda w.\transdn{e} t v (w)
&
\transdn{-e} t v &= \lambda w.\transup{e} t v (w)
\\
\transup{t} t v &= \lambda w.w + v 
&
\transdn{t} t v &= \lambda w.w + (\mathrm{negate}\, v)
\\
\transup{t'} t v &= \lambda w.w\quad(t'\neq t)
&
\transdn{t'} t v &= \lambda w.w\quad(t'\neq t)
\end{align*}

Unfolding definitions, we deduce that
\[
\tmSem{I(J(y))}(\opsEnv,y) = \tmSem{\transup A t v}(\opsEnv,\vec 0)(y)
\]
for any $y \in \tmSem{A[0/t]}$. It's easy to prove by induction on $A$
that $\tmSem{\transup A t v}(\opsEnv,\vec 0)(y) = y$, and so
$\ctxteq\DeltaWithB{y\mathord:Y}{J(I(y))}y Y$ follows.

For the opposite direction of the isomorphism, we appeal to 
\autoref{thm:soundness} and so it suffices to show
$\semeq{\DeltaWithB}{f\mathord:X}{J(I(f))}{f}{X}$; that
is, given $\rho\in\semSort\DeltaWithB$, for any $f$ and $f'$ with
$(f,f')\in\rsem X\rho$, it is the case that 
$(\tmSem{J(I(f))}(\opsEnv,f), f')\in\rsem X\rho$.
Expanding the premise, we have
\begin{eqnarray}\label{eqn:isopremise}
\forall \vec{t}\in \mathbb R ^ 2,
\forall \vec{v}\in \mathbb R ^ 2,
(f(\vec{v}),f'(\rho(B)\vec{v} + \vec{t}))\in\rsem A (\rho,\vec{t})
\end{eqnarray}
Expanding the conclusion, we have to show for any
$\vec{t_0}\in\mathbb R^2$ and $\vec{v_0}\in\mathbb R^2$, it is the case that
$(\tmSem{J(I(f))}(\opsEnv,f)(\vec{v_0}),f'(\rho(B)\vec{v_0} + \vec{t_0}))\in\rsem A (\rho,\vec{t_0})$.
Unfolding the definitions of $J$ and $I$ and expanding the erasure semantics, 
this amounts to showing that
\[
\forall \vec{t_0},\vec{v_0}\in \mathbb R ^ 2,
(\tmSem{\transup A t v}(\opsEnv,\vec{v_0})(f\ \vec 0),f'(\rho(B)\vec{v_0} + \vec{t_0}))\in\rsem A (\rho,\vec{t_0}).
\]
In order to make progress we prove first by induction on $A$ that for any $\vec v$ and $\vec w$, 
the meaning of $\transup A t v$ can be characterized by
\begin{align*}
(\tmSem{\transup A t v}(\opsEnv,\vec{v})(x),y)&\in\rsem A (\rho,\vec{w})
\\\Leftrightarrow 
(x,y)&\in\rsem A (\rho,\rho(B)\vec{v} + \vec{w})
\end{align*}
Thus our goal becomes
\[
\forall \vec{t_0},\vec{v_0}\in \mathbb R ^ 2,
(f\ \vec 0,f'(\rho(B)\vec{v_0} + \vec{t_0}))\in\rsem A (\rho,\rho(B)\vec{v_0} + \vec{t_0}).
\]
This is obtained immediately by instantiating \autoref{eqn:isopremise}
with $\vec{v}=\vec 0$ and $\vec{t}=\rho(B)\vec{v_0} + \vec{t_0}$.
\end{proof}

\exref{ex:type-iso} and \exref{ex:simpletrans} are special cases of
this isomorphism. Another instance is the type of vector addition:
\begin{align*}
\forall B\mathord:\SynGL{2}, t_1, t_2\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_2}\to\tyPrim{vec}{B,t_1\mathord+t_2}\\
\cong\quad\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_1}
\quad (\text{by~\lemref{lem:transiso}})\\
\cong\quad\forall B\mathord:\SynGL{2}.\tyPrim{vec}{B,0} \quad (\text{by~\lemref{lem:transiso}})\\
\cong\quad\tyUnit\quad(\text{by \autoref{thm:abstraction}})
\end{align*}
In other words, the vector addition operation is the only inhabitant of its type!
\paragraph{Scalings}
For the group $\GL{1}$ of scalings we can treat
a real-valued argument to a function as a unit-of-measure with
which to scale the result. 
%The small fly in the ointment is that the
Although the argument might be zero, and this cannot be used for scaling,
%Nevertheless, 
we still can obtain the following slightly more complicated
isomorphism:
\begin{eqnarray}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A &
\cong & [1/s]\times \forall s\mathord:\SynGL{1}.A\label{eqn:scaleiso}
\end{eqnarray}
The type of $\mathrm{areaCircle}$ from the introduction is one instance:
\begin{align*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s^2}\\ 
\cong\quad\tyPrim{real}{1}\times(\forall
s\mathord:\SynGL{1}.\tyPrim{real}{s^2}) \quad(\text{by \autoref{eqn:scaleiso}})\\ 
\cong\quad\tyPrim{real}{1}\times \tyUnit \quad (\text{by \autoref{thm:abstraction}})\\
\cong\quad\tyPrim{real}{1}\quad (\text{trivially})
\end{align*}
%where %the isomorphism 
%The abstraction theorem is used to prove
%$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s^2}\cong \tyUnit$.
% is proved using
%the abstraction theorem.
%\begin{enumerate}
%\item Two cases: when we have a total ``division'' operation, and when
%  we don't.
%\item Type isomorphisms, based on Smith Normal Form
% \item Non-definability results, based on the non-compositional
%   relational environments
% \item Alteration of the non-definability results in the light of
%   addition of a square root operation
% \item Mention the variation we can obtain by changing some types,
%   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
%\end{enumerate}
